Demes: a specification for interchanging complex demography

### 2021-06-19 paper.tex
First pass at main body.

\begin{itemize}
    \item Demographic models are comprised of one or more interacting
        populations, along with description, doi, time units and generation
        times.
    \item Populations (or demes) are treated as discrete collection of
        exchangeable individuals following a well-defined set of rules
    \item These rules include population sizes, times of existence, and other
        information about the mating system (e.g. selfing or cloning rates)
    \item Relationships between demes are defined by ancestors/descendents
        and possible continuous or instantaneous migration between coexisting
        demes
\end{itemize}

\subsection*{High-level model specification in YAML}

\begin{itemize}
    \item YAML allows us to write this information in a human-readable and
        structured manner (lists, keys/items, etc)
    \item Implicit values, so that models can be written compactly
    \item Example of an IM Model~\ref{code:im_model}
\end{itemize}

\begin{figure}
\begin{minipage}{0.48\textwidth}
\begin{tcolorbox}
\inputminted[fontsize=\scriptsize,linenos,numbersep=5pt]{yaml}{models/IM.yaml}
\end{tcolorbox}
\begin{tcolorbox}
\includegraphics[width=\textwidth]{fig/IM}
\end{tcolorbox}
\end{minipage}\hfill
\begin{minipage}{0.48\textwidth}
\begin{tcolorbox}
\inputminted[fontsize=\scriptsize,linenos,numbersep=5pt]{yaml}{models/IM-resolved.yaml}
\end{tcolorbox}
\end{minipage}\\
\captionof{listing}{An IM model.}
\label{code:im_model}
\end{figure}

@@

\begin{itemize}
    \item The Python API, which can read in a YAML file or be built using the
        Builder
    \item Demographic model validitation
    \item Unambiguous, redundant information
    \item Designed to be programatically read by simulation software backends
\end{itemize}

>>>>

The Demes specification is a formal data model for describing
the properties of populations over time,
along with some metadata and provenance information.
The data model is based on the ubiquitous JSON~\citep{bray2017javascript}
standard, and the structure formally defined using
JSON Schema~\citep{wright2020json}.
The Demes JSON schema rigorously defines the data model,
specifying the hierarchical structure and the properties that are allowed in different
contexts along with their types and permissable ranges. This schema, and its
accompanying documentation, fully describe the entities in the model and their
relationships, and the required behaviour of implementations. Since the
schema is definitive, we will not recapitulate the details
here, but instead focus on the high level properties of the model and
the rationale behind key design decisions.

In Demes, demographic models consist of one or more interacting
% JK: not sure at all about this, but let's see how it goes for now.
populations (or ``demes"; to avoid confusion with the name of the
specification itself we will use the term ``population'' in this
discussion, in the understanding that the terms are interchangeable
for our purposes). Populations are defined as a collection of
exchangable individuals that exist for some period of time,
following a well-defined set of rules regarding population
sizes, mating systems, etc. Ancestor/descendent relationships
between populations can be defined, as well as
possible continuous or instantaneous migration between coexisting
populations. The history of each population is defined as a
series of ``epochs'' during which the population's parameters are
fixed.

The design of Demes is a balance between two partially competing requirements:
that (a) models should be easy for humans to understand and manipulate; and (b)
software processing Demes models should be provided with an unambiguous
representation that is straightforward to process. For efficiency of
understanding and avoidance of model specification error, we require a data
representation without redundancy (i.e., repetition of values). But, for the
simplicity of software working with the Demes model (and the avoidance of
programming error, or divergence in interpretations of the specification)
it is preferable to have an explicit representation, in
which all relevant values are readily available. Thus, we consider
three entities: the high-level, human readable, model specification;
the low-level, fully-qualified model specification; and the parser,
which is responsible for transforming the former into the latter.

\begin{figure}
\begin{minipage}{0.48\textwidth}
\begin{tcolorbox}
\inputminted[fontsize=\scriptsize,linenos,numbersep=5pt]{yaml}{models/IM.yaml}
\end{tcolorbox}
\begin{tcolorbox}
\includegraphics[width=\textwidth]{fig/IM}
\end{tcolorbox}
\end{minipage}\hfill
\begin{minipage}{0.48\textwidth}
\begin{tcolorbox}
\inputminted[fontsize=\scriptsize,linenos,numbersep=5pt]{yaml}{models/IM-resolved.yaml}
\end{tcolorbox}
\end{minipage}\\
\caption{\label{fig-example}
Example isolation-with-migration Demes model. (A) The compact non-redundant
representation expressed using YAML. (B) The same model in the fully-qualified
form. (C) A visual representation of the model using \texttt{demesdraw}.
The YAML descriptions in (A) and (B) correspond exactly to a JSON description,
but are much more concise.
}
\end{figure}

\subsection*{High-level model specification}

The high-level description of Demes models is focused on the efficiency of
human understanding and the avoidance of errors. We have adopted the widely
used YAML format~\citep{ben2009yaml} as the recommended means of interchanging
Demes models (see Figure~\ref{fig-example} for exampels). YAML is a data
serialisation language with an emphasis on simplicity, and interoperates well
with JSON (indeed, YAML 1.2 is a superset of JSON). We chose YAML over JSON
because although JSON is an excellent format for data interchange, it is
ill-suited for human understanding or manipulation. We also considered other
data formats such as TOML, but chose YAML because of its greater popularity and
software support. Since the Demes data model is defined in JSON Schema,
however, there is no formal dependency on YAML and implementations may choose
to use JSON directly if they wish (e.g., for greater efficiency).

Structurally, the high-level Demes specification encourages human understanding
by avoiding redundancy in the description where possible, and
by providing a mechanism for specifying default values that
are inherited hierarchically. Figure~\ref{fig-example} shows an example
Demes model expressed in both the high and low-level forms (both in
terms of YAML syntax).

\jkcomment{Finish up this section with a discussion of how we (a)
use default values defined in JSON schema; (b) define more
subtle implicit values in the specification itself; and (c) use the
defaults sections, all explified by Fig 1.}

\subsection*{Parsers}
The implicit value inference and default value propagation required to
fully resolve the high-level Demes model description is straightforward
to describe, but still requires significant effort to implement in
software. Moreover, there are many constraints on the data model
(for example, epoch \texttt{end\_time} values should be strictly increasing
within a deme), which must be enforced.
It would not be reasonable to require every program that
takes Demes models as input to implement this logic, as the programming
effort would be significant (limiting adoption)
and it is likely that if there were many implementations some would differ
in their details (harming the software ecosystem). We therefore define
a standard software entity as part of the specification (the parser),
which performs this task, and which can be shared by programs that
support demes as an input. By having relatively few (ideally, one
per programming language), high-quality Demes parsers available as
libraries the probability of divergences from the standard is
greatly reduced.

The Demes specification precisely defines the required behaviour of
parsers, and we also provide a reference implementation written
in Python to resolve any potential ambiguities and to provide a
helpful template for other implementations. We also provide an
extensive suite suite of examples and the expected outputs. In addition,
we have high-quality parser implementations in the Python and C
languages (published under liberal open source licences),
providing a solid foundation for the software ecosystem.

@@

\jkcomment{Not sure if there's much point in this as a separate section any
more? This could really be said in the previous section.}

Parsers output the low-level description of Demes models,
which is an explicit representation
of the model where all parameters have been assigned values, and
all data constraints have been checked. The output is formally defined
as JSON, but in practise parsers output an object model that
is suitable for the particular programming language, and
that can be used directly by the implementing program.
